Visual Diagnostics for Constrained Optimisation with Application to Guided Tours

### 2021-01-31 paper/zhang-cook-laa-langrene-menendez.Rmd
add the plot for the orientation section
writing on comparing the three for kol_norm

## Flipping the sign in the bases

@@

The index function, up until this point, are all smooth, while this is not the case for all. A 1D projection function based on the Kolmogorov test, `norm_kol`, compares the projected data, $\mathbf{Y}_{n \times 1}$ to a randomly generated normal distribution, $y_n$ based on the empirical cumulated distribution function (ECDF). Let $F_{.}(u)$ be the ECDF function with the subscript $P$ and $u$ indicating the projected and randomly generated ECDF respectively,  the index is defined as:

$$\max \left[F_{P}(u) - F_{y}(u)\right]$$
With a more complex index function, it is interesting to understand 1) if all the three optimisers can reach the optimum, 2) when a local optimum is presented, whether the simulated annealing algorithm can escape that local optimum and find the global one.

```{r noisy-better-geo, fig.cap = "1D projection on 5\\-variable dataset \\code{boa5} with \\code{norm\\_kol} index. The trace plot shows the failure of pseudo derivative on non-smooth index. The PCA plot also shows that simulated annealing is searching the area close to the theoretical best while pseudo derivative can't couple with the non-continuity in the index."}
```

Figure \ref{fig:noisy-better-geo} presents the tracing plots of two optimisers: pseudo derivative and simulated annealing and as expected, the interpolated path is no longer smooth in either case. There is barely any improvement on the index value in the pseudo derivative algorithm while the simulated annealing algorithm has managed to get close to the index value of the theoretical best, indicated by the horizontal dashed line. In the PCA plot, simulated annealing is able to move gradually towards the theoretical best while the pseudo derivative method breaks after the initial few attempts.

>>>>

## Reconciling the orientation

```{r flip-sign, fig.cap = "PCA plot of 1D projection of boa6 data, optimised by SA, with seed 2463."}
```

@@

The index function before this section are all smooth, while this is not the case for all. A 1D projection function based on the Kolmogorov test, $I^{nk}(n)$, compares the projected data, $\mathbf{Y}_{n \times 1}$, to a randomly generated normal distribution, $\mathcal{N}_{n \times 1}$, based on the empirical cumulated distribution function (ECDF). Let $F_{.}(n)$ be the ECDF function, with two possible subscripts $Y$ and $\mathcal{N}$ representing the projected and randomly generated data, and $n$ denoting the number of observation, the Normal Kolmogorov, $I^{nk}(n)$, index is defined as:

$$I^{nk}(n) = \max \left[F_{Y}(n) - F_{\mathcal{N}}(n)\right]$$
With a non-smooth index function, two research questions are raised to understand

1) if any optimiser fails to optimise a non-smooth index; and 
2) whether the optimisers can find the global optimum despite the presence of a local optimum

```{r noisy-better-geo, fig.height = 6, fig.cap = "Trace and PCA plot of 1D projection on 5\\-variable dataset, \\code{boa5}, with \\code{norm\\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best."}
```

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of all three optimisers. As expected, none of the interpolated path is smooth. There is barely any improvement made by PD, indicating its failure in optimising non-smooth index. While SA and SAJO have managed to get close to the index value of the theoretical best, trace plot shows that it takes SAJO much longer to interpolate towards the final basis. This long interpolation path is partially due to the fluctuation in the early iterations, where the SAJO tends to generously accept inferior basis before concentrating near the optimum. The PCA plot shows the interpolation path and search points excluding the last termination iteration. [A sentence for PD]. Comparing the amount of random search done by SA and SAJO, it is surprising that although SAJO doesn't actually carry that many of search as SA. Combine the insight from both the trace and PCA plot, one can learn the two different search strategies by SA and SAJO: SA frequently samples in the space and only make a move when an improvement will be made, while SAJO first broadly accepts basis in the space and only starts the extensive search in a narrowed subspace. The specific decision of which optimiser to use will depends on the index curve in the basis space but if the basis space is highly non-smooth and spike, accepting inferior bases at first, as SAJO, can lead to a more efficient search, in terms of the overall number of points evaluated.

### 2021-01-31 paper/zhang-cook-laa-langrene-menendez.tex
writing on comparing the three for kol_norm

The index function, up until this point, are all smooth, while this is
not the case for all. A 1D projection function based on the Kolmogorov
test, \texttt{norm\_kol}, compares the projected data,
\(\mathbf{Y}_{n \times 1}\) to a randomly generated normal distribution,
\(y_n\) based on the empirical cumulated distribution function (ECDF).
Let \(F_{.}(u)\) be the ECDF function with the subscript \(P\) and \(u\)
indicating the projected and randomly generated ECDF respectively, the
index is defined as:

\[\max \left[F_{P}(u) - F_{y}(u)\right]\] With a more complex index
function, it is interesting to understand 1) if all the three optimisers
can reach the optimum, 2) when a local optimum is presented, whether the
simulated annealing algorithm can escape that local optimum and find the
global one.

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1} \caption[1D projection on 5\-variable dataset \code{boa5} with \code{norm\_kol} index]{1D projection on 5\-variable dataset \code{boa5} with \code{norm\_kol} index. The trace plot shows the failure of pseudo derivative on non-smooth index. The PCA plot also shows that simulated annealing is searching the area close to the theoretical best while pseudo derivative can't couple with the non-continuity in the index.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

Figure \ref{fig:noisy-better-geo} presents the tracing plots of two
optimisers: pseudo derivative and simulated annealing and as expected,
the interpolated path is no longer smooth in either case. There is
barely any improvement on the index value in the pseudo derivative
algorithm while the simulated annealing algorithm has managed to get
close to the index value of the theoretical best, indicated by the
horizontal dashed line. In the PCA plot, simulated annealing is able to
move gradually towards the theoretical best while the pseudo derivative
method breaks after the initial few attempts.

>>>>

The index function before this section are all smooth, while this is not
the case for all. A 1D projection function based on the Kolmogorov test,
\(I^{nk}(n)\), compares the projected data, \(\mathbf{Y}_{n \times 1}\),
to a randomly generated normal distribution,
\(\mathcal{N}_{n \times 1}\), based on the empirical cumulated
distribution function (ECDF). Let \(F_{.}(n)\) be the ECDF function,
with two possible subscripts \(Y\) and \(\mathcal{N}\) representing the
projected and randomly generated data, and \(n\) denoting the number of
observation, the Normal Kolmogorov, \(I^{nk}(n)\), index is defined as:

\[I^{nk}(n) = \max \left[F_{Y}(n) - F_{\mathcal{N}}(n)\right]\] With a
non-smooth index function, two research questions are raised to
understand

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  if any optimiser fails to optimise a non-smooth index; and
\item
  whether the optimisers can find the global optimum despite the
  presence of a local optimum
\end{enumerate}

\begin{Schunk}
\begin{figure}
\includegraphics[width=1\linewidth]{/Users/hzha400/Documents/PhD/research/paper-tour-vis/figs/noisy-better-geo-1} \caption[Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers]{Trace and PCA plot of 1D projection on 5\-variable dataset, \code{boa5}, with \code{norm\_kol} index, optimised by all three optimiers. Optimiser PD fails to optimise the non-smooth index while although both SA and SAJO success in optimise the index, SAJO takes much longer than SA and finishes off closer to the theoretial best.}\label{fig:noisy-better-geo}
\end{figure}
\end{Schunk}

Figure \ref{fig:noisy-better-geo} presents the trace and PCA plots of
all three optimisers. As expected, none of the interpolated path is
smooth. There is barely any improvement made by PD, indicating its
failure in optimising non-smooth index. While SA and SAJO have managed
to get close to the index value of the theoretical best, trace plot
shows that it takes SAJO much longer to interpolate towards the final
basis. This long interpolation path is partially due to the fluctuation
in the early iterations, where the SAJO tends to generously accept
inferior basis before concentrating near the optimum. The PCA plot shows
the interpolation path and search points excluding the last termination
iteration. {[}A sentence for PD{]}. Comparing the amount of random
search done by SA and SAJO, it is surprising that although SAJO doesn't
actually carry that many of search as SA. Combine the insight from both
the trace and PCA plot, one can learn the two different search
strategies by SA and SAJO: SA frequently samples in the space and only
make a move when an improvement will be made, while SAJO first broadly
accepts basis in the space and only starts the extensive search in a
narrowed subspace. The specific decision of which optimiser to use will
depends on the index curve in the basis space but if the basis space is
highly non-smooth and spike, accepting inferior bases at first, as SAJO,
can lead to a more efficient search, in terms of the overall number of
points evaluated.
